Helmet Foam Case Study Report

### 2024-04-27 CS2 Final Draft.qmd
Add files via upload

---
title: "Untitled"
output: html_document
date: "2024-04-24"
---

@@

@@

This project comes to us from Moria Foster, from the Lawrence Livermore National Laboratory. Foster has been running experiments on the Colorado School of Mines campus to study the fatigue responses of stiffness and damping in different varieties of foam. In her designed experiment, pieces of foam were put into a universal load frame, which repeatedly compressed and decompressed each piece for thousands of cycles, and each piece had its stiffness and damping measured periodically. These measurements, as well as the parameters of each experimental run, were given to us to aid in the experiment’s analysis stage.

@@

Concerns regarding the data's adequacy for drawing cause-and-effect conclusions prompted us to reformat the data from wide to long, helping our statistical analysis by ensuring each observation represents one measurement. Additionally, to maintain consistency across observations, samples taken beyond 10,000 cycles were removed. This step ensured that our analysis focuses on the relevant time frame for evaluating foam responses. By doing this we are hoping that our EDA gives us helpful insights into how we can optimize foam padding in these military helmets.

There were a few concerning observations with negative values of stiffness and/or damping. At Foster’s instruction, we converted the negative stiffness values to positive, and left the negative damping values how they were.

@@

```{r}
foam_Samples <- foam_data[ ,c("ID", "chemIndex", "porosity", "freq", "amp", "initialStrain", "initialStress")]
foam_Samples <- unique(foam_Samples)

The counts above show that unequal amounts of each foam type were tested. If a specific type of foam (like low-porosity foam with chemical index 2) has very few observations, it will be more difficult to make solid conclusions about this type of foam. Even the foam types with more samples still have relatively few, so making solid conclusions for these could be difficult as well.

```{r}
#| echo: false
#| label: fig-1
#| warning: false
#| fig-cap: "Histogram of Frequencies Tested"
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 4
#| 
ggplot(foam_Samples, aes(x=freq)) +
  geom_histogram()
```

@@

```{r}
#| echo: false
#| label: fig-2
#| warning: false
#| fig-cap: "siffnessFrac vs cycle"
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 4

@@

p2 <- ggplot(foam_data_scaled, aes(y = dampingFrac, x = cycle, col=factor(ID))) + 
  geom_point() +
  scale_y_continuous(limits = c(0.825, 1)) + 
  theme(legend.position="none")

p3 <- ggplot(foam_data_scaled, aes(y = dampingFrac, x = cycle, col=factor(ID))) + 
  geom_point() +
  scale_y_continuous(limits = c(-20, 1)) +
  theme(legend.position="none")

combined <- p1 + p2 + p3

combined

```
In order to see how stiffnessFrac and dampingFrac change over time for all foams, we plotted a scatter plot for each vs cycle. The dot colors in @fig-2 represent which foam sample they were measured from. In the first plot, stiffnessFrac seems to spread out in an umbrella shape, showing that as cycle increases the sample values not only change, but get further apart. So cycle number has a relationship with stiffnessFrac, but that relationship looks different for each foam sample.

@@

The second plot was scaled so that the scale was from 0.75-1, given that all but 13 dampingFrac observations were in this range. The observations outside of this range (included in the unscaled third plot) all occurred in the interval code N, which is the last interval code we have in our filtered data, most likely meaning that these foams broke in some way or the experiment failed.

@@

In @fig-3, we logged the responses, and plotted them against the log of cycle. The relationships between log(cycle) and the logged responses look more linear than the relationships in @fig-2.

In order to understand how log(stiffnessFrac) and log(dampingFrac) behave over log(cycle) given different frequencies, we also colored the samples in @fig-3 by frequency. For log(stiffnessFrac), we see that higher frequencies have a more negative trend than lower frequencies. For log(dampingFrac), we see that lower frequencies have a more negative trend than higher frequencies. So frequency looks related to both stiffnessFrac and dampingFrac.

```{r}
#| echo: false
#| label: fig-4
#| warning: false
#| fig-cap: "log(stiffnessFrac) vs log(cycle) by chemIndex and porosity"
#| fig-align: "center"
#| fig-width: 18
#| fig-height: 8
#| 
sfc1 <- ggplot(foam_data_scaled[foam_data_scaled$chemIndex == 1 & foam_data$porosity == "high",], aes(x=log(cycle), y=log(stiffnessFrac))) +
  geom_point(color="red", size=3) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  ylim(-0.325, 0.15)
sfc2 <- ggplot(foam_data_scaled[foam_data_scaled$chemIndex == 2 & foam_data$porosity == "high",], aes(x=log(cycle), y=log(stiffnessFrac))) +
  geom_point(color="green", size=3) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  ylim(-0.325, 0.15)
sfc3 <- ggplot(foam_data_scaled[foam_data_scaled$chemIndex == 3 & foam_data$porosity == "high",], aes(x=log(cycle), y=log(stiffnessFrac))) +
  geom_point(color="cyan2", size=3) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  ylim(-0.325, 0.15)
sfc4 <- ggplot(foam_data_scaled[foam_data_scaled$chemIndex == 1 & foam_data$porosity == "low",], aes(x=log(cycle), y=log(stiffnessFrac))) +
  geom_point(color="red3", size=3) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  ylim(-0.325, 0.15)
sfc5 <- ggplot(foam_data_scaled[foam_data_scaled$chemIndex == 2 & foam_data$porosity == "low",], aes(x=log(cycle), y=log(stiffnessFrac))) +
  geom_point(color="green4", size=3) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  ylim(-0.325, 0.15)
sfc6 <- ggplot(foam_data_scaled[foam_data_scaled$chemIndex == 3 & foam_data$porosity == "low",], aes(x=log(cycle), y=log(stiffnessFrac))) +
  geom_point(col="blue", size=3) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  ylim(-0.325, 0.15)

@@

The plots in @fig-4 separate the log(stiffnessFrac) vs log(cycle) graph from above to show how the interaction between chemical index and porosity impact log(stiffnessFrac). The top row shows only high porosity and the bottom row shows only low porosity samples. From left to right, the columns show chemical index 79, 100 and 121. We can see that the chemical index 79 has many samples where log(stiffnessFrac) trends upwards, whereas chemical indices 100 and 121 mostly trend downwards. This indicates a relationship between log(stiffnessFrac) and chemical index. Downward trends mean the stiffness decreases as cycle increases, which is not what we are looking for in a durable foam.

@@

```{r}
#| echo: false
#| label: fig-5
#| warning: false
#| fig-cap: "log(dampingFrac) vs log(cycle) for chemIndex and porosity"
#| fig-align: "center"
#| fig-width: 18
#| fig-height: 8
#| 
dfc1 <- ggplot(foam_data_scaled[foam_data_scaled$chemIndex == 1 & foam_data_scaled$porosity == "high",], aes(x=log(cycle), y=log(dampingFrac))) +
  geom_point(color="red", size=3) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  ylim(-0.175, 0.125)
dfc2 <- ggplot(foam_data_scaled[foam_data_scaled$chemIndex == 2 & foam_data_scaled$porosity == "high",], aes(x=log(cycle), y=log(dampingFrac))) +
  geom_point(color="green", size=3) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  ylim(-0.175, 0.125)
dfc3 <- ggplot(foam_data_scaled[foam_data_scaled$chemIndex == 3 & foam_data_scaled$porosity == "high",], aes(x=log(cycle), y=log(dampingFrac))) +
  geom_point(color="cyan2", size=3) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  ylim(-0.175, 0.125)
dfc4 <- ggplot(foam_data_scaled[foam_data_scaled$chemIndex == 1 & foam_data_scaled$porosity == "low",], aes(x=log(cycle), y=log(dampingFrac))) +
  geom_point(color="red3", size=3) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  ylim(-0.175, 0.125)
dfc5 <- ggplot(foam_data_scaled[foam_data_scaled$chemIndex == 2 & foam_data_scaled$porosity == "low",], aes(x=log(cycle), y=log(dampingFrac))) +
  geom_point(color="green4", size=3) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  ylim(-0.175, 0.125)
dfc6 <- ggplot(foam_data_scaled[foam_data_scaled$chemIndex == 3 & foam_data_scaled$porosity == "low",], aes(x=log(cycle), y=log(dampingFrac))) +
  geom_point(col="blue", size=3) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  ylim(-0.175, 0.125)

@@

The plots in @fig-5 separate the log(dampingFrac) vs log(cycle) graph from above to show how the interaction between chemical index and porosity impact log(dampingFrac). The data shown in the rows and columns are the same as in @fig-4. We can see differences in the log(dampingFrac) trends between different chemical indices, so there is likely a relationship present, but this relationship looks less pronounced than it was for the log(stiffnessFrac) trends.

@@

@@

>>>>

@@

foam_data_scaled <- foam_data |>
  filter(cycle <= 10000 & cycle > 0) |>
  mutate(dampingFrac = replace(dampingFrac, dampingFrac < 0, NA))

@@

This project comes to us from Moira Foster, from the Lawrence Livermore National Laboratory. Foster has been running experiments on the Colorado School of Mines campus to study the fatigue responses of stiffness and damping in different varieties of foam. In her designed experiment, pieces of foam were put into a universal load frame, which repeatedly compressed and decompressed each piece for thousands of cycles, and each piece had its stiffness and damping measured periodically. These measurements, as well as the parameters of each experimental run, were given to us to aid in the experiment’s analysis stage.

@@

We reformatted the data from wide to long to ensure each observation represents one measurement, not one foam sample. Measurements taken beyond 10,000 cycles were removed to maintain consistency across observations, and because Moira identified this as the most relevant time frame.

There were a few concerning observations with negative values of stiffness, likely due to a fault in the measuring machine. At Foster’s instruction, we converted these negative stiffness values to positive. The transformed response dampingFrac also had five concerning observations, that were not just negative, but drastic outliers. These were changed to NA per Dr. Philtron's recommendation.

@@

| Porosity | **ChemIndex 79** | **ChemIndex 100** | **ChemIndex 121** |
|---------:|:-----------------|:------------------|:------------------|
| **81**   | 9                | 7                 | 6                 |
| **71**   | 11               | 5                 | 8                 |

These counts show that a relatively small number of each type of foam were tested. The amounts are unequal as well, ranging from 5 to 11.

```{r}
#| echo: false
#| label: fig-1
#| warning: false
#| fig-cap: "Histogram of Frequencies Tested"
#| fig-align: "center"
#| fig-width: 4
#| fig-height: 4

foam_Samples <- foam_data[ ,c("ID", "chemIndex", "porosity", "freq", "amp", "initialStrain", "initialStress")]
foam_Samples <- unique(foam_Samples)

ggplot(foam_Samples, aes(x=freq)) +
  geom_histogram(bins=10)
```

@@

```{r}
#| echo: false
#| label: fig-2
#| warning: false
#| fig-cap: "StiffnessFrac and DampingFrac vs Cycle"
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 4

@@

p2 <- ggplot(foam_data_scaled, aes(y = dampingFrac, x = cycle, col=factor(ID))) + 
  geom_point() +
  theme(legend.position="none")

combined <- p1 + p2

combined
```
In order to see how stiffnessFrac and dampingFrac change over time for all foams, we plotted a scatter plot for each vs cycle. The dot colors in @fig-2 represent which foam sample they were measured from. As cycle number increases, the stiffnessFrac and dampingFrac values not only change, but spread further apart. So stiffnessFrac and dampingFrac both appear to have a relationship with cycle, but that relationship looks different for each foam sample.

@@

@@

In @fig-3, we logged the responses, and plotted them against the log of cycle. The relationships between log(cycle) and the logged responses look more linear than the relationships in @fig-2, so a linear model using these logged variables would fit better.

In order to understand how log(stiffnessFrac) and log(dampingFrac) behave given different frequencies, we also colored the samples in @fig-3 by frequency. For log(stiffnessFrac), we see that higher frequencies have a more negative trend than lower frequencies. For log(dampingFrac), we see that lower frequencies have a more negative trend than higher frequencies. So frequency looks related to both stiffnessFrac and dampingFrac.

```{r}
#| echo: false
#| label: fig-4
#| warning: false
#| fig-cap: "log(stiffnessFrac) vs log(cycle) by chemIndex and porosity"
#| fig-align: "center"
#| fig-width: 18
#| fig-height: 8
#| 
sfc1 <- ggplot(foam_data_scaled[foam_data_scaled$chemIndex == 1 & foam_data$porosity == "high",], aes(x=log(cycle), y=log(stiffnessFrac))) +
  geom_point(color="red", size=3) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  ylim(-0.325, 0.15) +
  labs(title="ChemIndex 79, Porisity 81") +
  theme(plot.title = element_text(size=25))
sfc2 <- ggplot(foam_data_scaled[foam_data_scaled$chemIndex == 2 & foam_data$porosity == "high",], aes(x=log(cycle), y=log(stiffnessFrac))) +
  geom_point(color="green", size=3) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  ylim(-0.325, 0.15) +
  labs(title="ChemIndex 100, Porisity 81") +
  theme(plot.title = element_text(size=25))
sfc3 <- ggplot(foam_data_scaled[foam_data_scaled$chemIndex == 3 & foam_data$porosity == "high",], aes(x=log(cycle), y=log(stiffnessFrac))) +
  geom_point(color="cyan2", size=3) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  ylim(-0.325, 0.15) +
  labs(title="ChemIndex 121, Porisity 81") +
  theme(plot.title = element_text(size=25))
sfc4 <- ggplot(foam_data_scaled[foam_data_scaled$chemIndex == 1 & foam_data$porosity == "low",], aes(x=log(cycle), y=log(stiffnessFrac))) +
  geom_point(color="red3", size=3) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  ylim(-0.325, 0.15) +
  labs(title="ChemIndex 79, Porisity 71") +
  theme(plot.title = element_text(size=25))
sfc5 <- ggplot(foam_data_scaled[foam_data_scaled$chemIndex == 2 & foam_data$porosity == "low",], aes(x=log(cycle), y=log(stiffnessFrac))) +
  geom_point(color="green4", size=3) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  ylim(-0.325, 0.15) +
  labs(title="ChemIndex 100, Porisity 71") +
  theme(plot.title = element_text(size=25))
sfc6 <- ggplot(foam_data_scaled[foam_data_scaled$chemIndex == 3 & foam_data$porosity == "low",], aes(x=log(cycle), y=log(stiffnessFrac))) +
  geom_point(col="blue", size=3) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  ylim(-0.325, 0.15) +
  labs(title="ChemIndex 121, Porisity 71") +
  theme(plot.title = element_text(size=25))

@@

The plots in @fig-4 separate the log(stiffnessFrac) vs log(cycle) graph in @fig-3 to show how the interaction between chemical index and porosity impact log(stiffnessFrac). We can see that the chemical index 79 has many samples where log(stiffnessFrac) trends upwards, whereas chemical indices 100 and 121 mostly trend downwards. This indicates a relationship between log(stiffnessFrac) and chemical index. Downward trends mean the stiffness decreases as cycle increases, which is not what we are looking for in a durable foam.

@@

```{r}
#| echo: false
#| label: fig-5
#| warning: false
#| fig-cap: "log(dampingFrac) vs log(cycle) for chemIndex and porosity"
#| fig-align: "center"
#| fig-width: 18
#| fig-height: 8
#| 
dfc1 <- ggplot(foam_data_scaled[foam_data_scaled$chemIndex == 1 & foam_data_scaled$porosity == "high",], aes(x=log(cycle), y=log(dampingFrac))) +
  geom_point(color="red", size=3) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  ylim(-0.175, 0.125) +
  labs(title="ChemIndex 79, Porisity 81") +
  theme(plot.title = element_text(size=25))
dfc2 <- ggplot(foam_data_scaled[foam_data_scaled$chemIndex == 2 & foam_data_scaled$porosity == "high",], aes(x=log(cycle), y=log(dampingFrac))) +
  geom_point(color="green", size=3) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  ylim(-0.175, 0.125) +
  labs(title="ChemIndex 100, Porisity 81") +
  theme(plot.title = element_text(size=25))
dfc3 <- ggplot(foam_data_scaled[foam_data_scaled$chemIndex == 3 & foam_data_scaled$porosity == "high",], aes(x=log(cycle), y=log(dampingFrac))) +
  geom_point(color="cyan2", size=3) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  ylim(-0.175, 0.125) +
  labs(title="ChemIndex 121, Porisity 81") +
  theme(plot.title = element_text(size=25))
dfc4 <- ggplot(foam_data_scaled[foam_data_scaled$chemIndex == 1 & foam_data_scaled$porosity == "low",], aes(x=log(cycle), y=log(dampingFrac))) +
  geom_point(color="red3", size=3) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  ylim(-0.175, 0.125) +
  labs(title="ChemIndex 79, Porisity 71") +
  theme(plot.title = element_text(size=25))
dfc5 <- ggplot(foam_data_scaled[foam_data_scaled$chemIndex == 2 & foam_data_scaled$porosity == "low",], aes(x=log(cycle), y=log(dampingFrac))) +
  geom_point(color="green4", size=3) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  ylim(-0.175, 0.125) +
  labs(title="ChemIndex 100, Porisity 71") +
  theme(plot.title = element_text(size=25))
dfc6 <- ggplot(foam_data_scaled[foam_data_scaled$chemIndex == 3 & foam_data_scaled$porosity == "low",], aes(x=log(cycle), y=log(dampingFrac))) +
  geom_point(col="blue", size=3) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  ylim(-0.175, 0.125) +
  labs(title="ChemIndex 121, Porisity 71") +
  theme(plot.title = element_text(size=25))

@@

The plots in @fig-5 separate the log(dampingFrac) vs log(cycle) graph from above to show how the interaction between chemical index and porosity impact log(dampingFrac). We can see differences in the log(dampingFrac) trends between different chemical indices, so there is likely a relationship present, but this relationship looks less pronounced than it was for the log(stiffnessFrac) trends.

@@

This is especially true for foams with especially few samples, like low-porosity foam with chemical index 100. Our model estimates regarding foams like this are even more liable to be wrong. So our models are not equally strong with all 6 types of foam.

@@

```{r, include=TRUE}
foam_Samples <- foam_data[ ,c("ID", "chemIndex", "porosity", "freq", "amp", "initialStrain", "initialStress")]
foam_Samples <- unique(foam_Samples)
